---
title: "SHADE Workflow Demo with Simulated Spatial Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

This demo illustrates the SHADE model described in the paper "SHADE: Spatial Hierarchical Asymmetry via Directional Estimation." SHADE is a Bayesian hierarchical model designed to estimate asymmetric spatial associations between cell types across spatial scales and biological levels (image, patient, cohort).

We will:
- Simulate spatial point patterns with biologically plausible asymmetry
- Construct interaction features using radial basis functions
- Prepare data for Bayesian inference using logistic approximation
- Fit the SHADE model
- Summarize and compare posterior results to ground truth

## 1. Load Packages

We start by loading all necessary libraries, including your SHADE package and its dependencies.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(SHADE)
library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(Matrix)
library(spatstat.geom)
library(spatstat.random)
library(posterior)
```

## 2. Set Simulation Parameters

We simulate a small dataset of tissue images with three cell types. One of these is treated as the "target" cell type whose spatial intensity is influenced by the presence of others.

```{r}
sim_idx <- 1
grid <- expand.grid(num_pts_per_group = c(10, 20, 40),
                    images_per_pt = c(1, 2, 4),
                    sim = 1:15)

num_pts_per_group <- grid$num_pts_per_group[sim_idx]
images_per_pt     <- grid$images_per_pt[sim_idx]
sim               <- grid$sim[sim_idx]

num_types <- 3
np <- 150
ratio <- 2
n_dummy <- floor(np * ratio)
num_pot <- 3
potentials <- make_rbfs(n_basis_functions = num_pot, max_dist = 75, basis_function_sigma = 15)
mean_alpha <- log(np / (1500^2))

params <- make_simulation_parameters(
  mean_alpha = mean_alpha,
  sigma_beta_global = 0.5,
  sigma_beta_indiv  = 0.1,
  sigma_beta_local  = 0.1,
  scale_sigmas = 5,
  num_pt_groups = 1,
  num_types = num_types,
  num_combos = num_types - 1,
  num_pot = num_pot,
  indiv_to_group = rep(1, num_pts_per_group),
  num_pts = num_pts_per_group,
  num_images = num_pts_per_group * images_per_pt,
  sample_to_indiv = rep(1:num_pts_per_group, each = images_per_pt),
  seed = 2024 + sim
)
betas_local <- params$betas_local[-1, ]
```

## 3. Simulate Spatial Point Patterns

We simulate images where target cell placement depends on background cell types. Conditional intensities are generated using spatial interaction fields (SIFs), which are sums of RBFs weighted by coefficients.

```{r}
W <- owin(c(0, 1500), c(0, 1500))
area <- 1500^2
num_images <- num_pts_per_group * images_per_pt

sim_data <- lapply(1:num_images, function(i) {
  if (num_types == 2) {
    pat <- rpoispp(lambda = np / area, win = W)
    marks(pat) <- factor("t1")
  } else {
    pat <- rmpoispp(lambda = rep(np / area, num_types - 1), win = W)
  }

  dens_list <- lapply(1:(num_types - 1), function(j) {
    coeffs <- betas_local[((j - 1) * num_pot + 1):(j * num_pot), i]
    custom_kernel <- Vectorize(function(x, y) {
      d <- sqrt(x^2 + y^2)
      sum(sapply(seq_along(coeffs), function(k) coeffs[k] * potentials[[k]](d)))
    })
    subs <- unmark(subset(pat, marks == j))
    smooth_density_fft(subs, custom_kernel, resolution = 128)
  })

  interaction_density <- Reduce("+", dens_list)
  lambda_integral <- sum(exp(interaction_density$v)) * (interaction_density$xstep * interaction_density$ystep)
  beta0 <- log(np / lambda_integral)
  target_pat <- rpoispp(lambda = exp(interaction_density + beta0))
  marks(target_pat) <- factor(num_types)

  final_pat <- superimpose(pat, target_pat)
  list(pat = final_pat, beta0 = beta0)
})
```

## 4. Prepare the Data for SHADE

We extract spatial coordinates and cell type labels, and format metadata for hierarchical modeling. SHADE models variation across images, patients, and groups.

```{r}
coords <- do.call(rbind, lapply(seq_along(sim_data), function(i) {
  pat <- sim_data[[i]]$pat
  raw_marks <- as.character(pat$marks)
  relabeled_marks <- paste0("t", raw_marks)

  tibble(
    x = pat$x,
    y = pat$y,
    type = factor(relabeled_marks, levels = paste0("t", 1:num_types)),
    image_id = paste0("img_", i)
  )
}))

image_ids <- unique(coords$image_id)
patient_ids <- rep(paste0("pt_", 1:num_pts_per_group), each = images_per_pt)
patient_metadata <- tibble(
  Spot = image_ids,
  Patient = patient_ids,
  Group = rep("group_1", length(image_ids))
)

prep <- prepare_spatial_model_data(
  x = coords$x,
  y = coords$y,
  cell_type = coords$type,
  image_id = coords$image_id,
  patient_metadata = patient_metadata,
  type_idx = num_types,
  n_dummy = n_dummy,
  n_basis_functions = num_pot
)
```

## 5. Fit the SHADE Model

We use a logistic regression approximation to fit the conditional intensity function. Inference is performed via HMC using cmdstanr.

```{r}
fit <- run_SHADE_model(
  prep$stan_data,
  chains = 1,
  iter_warmup = 300,
  iter_sampling = 100,
  refresh = 10
)
```

## 6. Summarize Posterior Estimates

We summarize the posterior draws for global-level coefficients using the rvar object. This captures uncertainty in spatial association curves (SICs).

```{r}
rvars <- posterior::as_draws_rvars(fit$draws())
beta_global_rvar <- rvars$beta_global[-1, ]

posterior_global_tbl <- posterior::summarise_draws(
  beta_global_rvar, 
  mean, median, sd, mad,
  `5%` = ~quantile(.x, 0.05),
  `95%` = ~quantile(.x, 0.95),
  rhat, ess_bulk
) %>%
  filter(str_detect(variable, "\\[")) %>%
  separate(variable, into = c("param", "row", "col"), sep = "\\[|,|\\]", convert = TRUE) %>%
  mutate(term = paste0("β_global[", row, "] (Group ", col, ")"))

posterior_global_tbl %>%
  select(term, mean, sd, `5%`, `95%`, rhat, ess_bulk) %>%
  arrange(term)
```

## 7. Compare to Ground Truth Coefficients

We validate inference by comparing posterior estimates to the known coefficients used to generate the spatial interaction curves.

```{r}
params$betas_global[-1, ]
```

## Next Steps

- Visualize spatial interaction curves (SICs) across groups  
- Perform posterior predictive checks  
- Apply SHADE to real multiplexed imaging data  